---
output: github_document
bibliography: "inst/REFERENCES.bib"
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# R/`cmldiffvar`

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
[![MIT license](http://img.shields.io/badge/license-MIT-brightgreen.svg)](https://opensource.org/license/mit/)
[![R-CMD-check](https://github.com/PhilBoileau/cmldiffvar/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/PhilBoileau/cmldiffvar/actions/workflows/R-CMD-check.yaml)
[![Codecov test coverage](https://codecov.io/gh/PhilBoileau/cmldiffvar/graph/badge.svg)](https://app.codecov.io/gh/PhilBoileau/cmldiffvar)
<!-- badges: end -->


> Causal Machine Learning Methods for Differential Variance Inference

__Authors:__ [Philippe Boileau](https://pboileau.ca),
[Hani Zaki](https://ca.linkedin.com/in/hani-zaki),
[Mireille Schnizter](https://www.mireilleschnitzer.com)

---

## What's `cmldiffvar`?

`cmldiffvar` implements causal machine learning methods for differential
variance inference. These methods rely on semiparametric efficiency theory and
flexible machine learning methods --- namely, Super Learner ensembles --- to
avoid the need for convenience assumptions about data-generating processes
[@vanderlaanTargetedLearningCausal2011; @laanSuperLearner2007]. Hypothesis
tests about differential variance can uncover heterogeneous treatment effects,
even when the effect modifiers are excluded from the data.

---

## Installation

The _development version_ of the package may be installed from GitHub using
[`remotes`](https://CRAN.R-project.org/package=remotes):

```{r gh-main-installation, eval=FALSE}
remotes::install_github("PhilBoileau/cmldiffvar")
```

---

## Example

We estimate the absolute differential variance, defined as the difference of the
potential outcomes' standard deviations, on a random sample of the
`toy_population_tbl` data included with the `cmldiffvar` package. This dataset
represents an observational study in which the treatment variable is binary, the
outcome is continuous, and a single confounder was measured. The true absolute
differential variance in this population is $2$. Because the absolute
differential variance is non-zero, the treatment effect is heterogeneous.

We use a targeted maximum likelihood estimator, the `cmldiffvar()` function's
default estimator, to infer the differential variance of this population. The
function outputs a point estimate and a $95\%$ confidence interval by default. A
p-value corresponding to a test of whether the differential variance is
significantly different from zero is also provided. This is equivalent to
testing whether the treatment effect is homogeneous.

```{r example, message=FALSE, warning=FALSE}
# load the required packages
library(cmldiffvar)
library(dplyr)
library(SuperLearner)

# set the seed for reproducibility
set.seed(510)

# random sample from population data
sample_tbl <- slice_sample(toy_population_tbl, n = 250)

# estimate absolute differential variance
dif_var_result_tbl <- sample_tbl |>
  cmldiffvar(
    propensity_score_adj_var_names = "confounder",
    cond_exp_outcome_adj_var_names = "confounder",
    treatment_var_name = "treatment",
    outcome_var_name = "outcome"
  )
```
```{r print-table, echo=FALSE}
# print formatted table
library(knitr)
dif_var_result_tbl |> kable(digits = 2, format.args = list(nsmall = 2))
```

The absolute differential variance point estimate is near the ground truth.
Additionally, the test correctly rejects the null hypothesis of a homogeneous
treatment effect at the $5\%$ signigicance level.

---

## Issues

If you encounter any bugs or have any specific feature requests, please [file
an issue](https://github.com/PhilBoileau/cmldiffvar/issues).

---

## Contributions

Contributions are very welcome. Interested contributors should consult our
[contribution
guidelines](https://github.com/PhilBoileau/cmldiffvar/blob/main/.github/CONTRIBUTING.md)
prior to submitting a pull request.

---

## Citation

Please cite the following paper when using the `cmldiffvar` R software package.

```
@article{boileau-cmldiffvar,
  author = {Philippe boileau and Hani Zaki and Mireille Schnizter},
  journal = {To appear},
  title = {Causal Machine Learning Methods for Differential Varance Inference},
  url = {NA},
  year = {2025}
}
```

---

## Licence

&copy; 2025 [Philippe Boileau](https://pboileau.ca)

The contents of this repository are distributed under the MIT license. See file
[`LICENSE.md`](https://github.com/PhilBoileau/cmldiffvar/blob/main/LICENSE.md)
for details.

---

## References
