---
output: github_document
bibliography: "inst/REFERENCES.bib"
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# R/`cmldiffvar`

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
[![MIT license](http://img.shields.io/badge/license-MIT-brightgreen.svg)](https://opensource.org/license/mit/)
[![R-CMD-check](https://github.com/PhilBoileau/cmldiffvar/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/PhilBoileau/cmldiffvar/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->


> Causal Machine Learning Methods for Differential Variance Inference

__Authors:__ [Philippe Boileau](https://pboileau.ca),
[Hani Zaki](link-to-profile),
[Mireille Schnizter](https://www.mireilleschnitzer.com)

---

## What's `cmldiffvar`?

`cmldiffvar` implements causal machine learning methods for differential
variance inference. These methods rely on semiparametric efficiency theory and
flexible machine learning methods --- namely, Super Learner ensembles --- to
avoid the need for convenience assumptions about data-generating processes
[@vanderlaanTargetedLearningCausal2011; @laanSuperLearner2007].

---

## Installation

The _development version_ of the package may be installed from GitHub using
[`remotes`](https://CRAN.R-project.org/package=remotes):

```{r gh-main-installation, eval=FALSE}
remotes::install_github("PhilBoileau/cmldiffvar")
```

---

## Example

We estimate the absolute differential variance (ADV) using a cross-fit targeted
maximum likelihood estimator from a random sample of the `toy_population_tbl`
data included with the `cmldiffvar` package. The true ADV in this population is $8$.

```{r example, message=FALSE, warning=FALSE}
# load the required packages
library(cmldiffvar)
library(dplyr)
library(SuperLearner)
library(knitr)

# set the seed for reproducibility
set.seed(514)

# random sample from population data
sample_tbl <- slice_sample(toy_population_tbl, n = 500)

# estimate absolute differential variance
dif_var_result_tbl <- sample_tbl |>
  cmldiffvar(
    confounder_var_names = "confounder",
    treatment_var_name = "treatment",
    outcome_var_name = "outcome"
  )

# print formatted table
dif_var_result_tbl |> kable(digits = 2, format.args = list(nsmall = 2))
```

---

## Issues

If you encounter any bugs or have any specific feature requests, please [file
an issue](https://github.com/PhilBoileau/cmldiffvar/issues).

---

## Contributions

Contributions are very welcome. Interested contributors should consult our
[contribution
guidelines](https://github.com/PhilBoileau/cmldiffvar/blob/main/.github/CONTRIBUTING.md)
prior to submitting a pull request.

---

## Citation

Please cite the following paper when using the `cmldiffvar` R software package.

```
@article{boileau-cmldiffvar,
  author = {Philippe boileau and Hani Zaki and Mireille Schnizter},
  journal = {arXiv preprint},
  title = {Causal Machine Learning Methods for Differential Varance Inference},
  url = {NA},
  year = {2025}
}
```

---

## Licence

&copy; 2025 [Philippe Boileau](https://pboileau.ca)

The contents of this repository are distributed under the MIT license. See file
[`LICENSE.md`](https://github.com/PhilBoileau/cmldiffvar/blob/main/LICENSE.md)
for details.

---

## References
