---
title: "Using the cmldiffvar Package"
author: "[Philippe Boileau](https://pboileau.ca)"
date: "`r Sys.Date()`"
bibliography: ../inst/references.bib
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{using-cmldiffvar}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Background

The average treatment effect (ATE) is a common causal estimand in causal
analyses. Defined as the difference of the mean potential outcomes, it
corresponds to the difference in average outcomes had all study units been
assigned to the treatment groups versus had they all been assigned to the
control group. While the ATE provides an important summary of the treatment
effect, it says nothing of effect heterogeneity. Even when the ATE indicates a
net positive effect of treatment on the average outcome, some study units may
experience better outcomes when assigned to the control. Decisions informed
solely by inference about the ATE might therefore produce unwelcome results;
effect heterogeneity, if it exists, must be accounted for.

The conditional ATE (CATE), another causal parameter, has gained popularity as
it can uncover and quantify this effect heterogeneity. As its name suggests, the
CATE reports the ATE of study units as a function of their pre-treatment
covariates. However, the CATE can only capture treatment effect heterogeneity if
the considered covariates are treatment effect modifiers. When the true effect
modifiers are unknown or they are missing from the data due to, for example, 
resource constraints or measurement issues, then the CATE will incorrectly
suggest that a treatment effect is homogeneous. Another inferential strategy
not subject to these limitations must supplement inference about the ATE.

We propose causal inference about the potential outcomes' variances to
accomplish just that. In particular, we have shown that tests about contrasts of
potential outcomes' variances, which we refer to differential variance, can be
used to formally test the homogeneous treatment effect assumption
[@boileau2025]. We briefly intoduce this procedure in the next section.


# Differential Variance for Heterogeneous Treatment Effect Detection


## Absolute Differential Variance


## Relative Differential Variance


# `cmldiffvar` in Action

```{r setup}
library(cmldiffvar)
```


## Unadjusted Estimators


### Estimators


### Example


## Adjusted Estimators


### Estimators


### Example


# Session Information

```{r session-information, echo=FALSE}
sessionInfo()
```

# References
