---
title: "Using the cmldiffvar Package"
author: "[Philippe Boileau](https://pboileau.ca)"
date: "`r Sys.Date()`"
bibliography: ../inst/references.bib
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{using-cmldiffvar}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Background

The average treatment effect (ATE) is a common causal estimand in causal
analyses. Defined as the difference of the mean potential outcomes, it
corresponds to the difference in average outcomes had all study units been
assigned to the treatment groups versus had they all been assigned to the
control group. While the ATE provides an important summary of the treatment
effect, it says nothing of effect heterogeneity. Even when the ATE indicates a
net positive effect of treatment on the average outcome, some study units may
experience better outcomes when assigned to the control. Decisions informed
solely by inference about the ATE might therefore produce unwelcome results;
effect heterogeneity, if it exists, must be accounted for.

The conditional ATE (CATE), another causal parameter, has gained popularity as
it can uncover and quantify this effect heterogeneity. As its name suggests, the
CATE reports the ATE of study units as a function of their pre-treatment
covariates. However, the CATE can only capture treatment effect heterogeneity if
the considered covariates are treatment effect modifiers. When the true effect
modifiers are unknown or they are missing from the data due to, for example, 
resource constraints or measurement issues, then the CATE will incorrectly
suggest that a treatment effect is homogeneous. Another inferential strategy
not subject to these limitations must supplement inference about the ATE.

We propose causal inference about the potential outcomes' variances to
accomplish just that. In particular, we have shown that tests about contrasts of
potential outcomes' variances, which we refer to differential variance, can be
used to formally test the homogeneous treatment effect assumption
[@boileau2025]. We briefly introduce this procedure in the next section.


# Differential Variance Inference

## Complete Data Model

We assume that data are generated according to a parallel group study design
wherein study units are randomly assigned to one of two groups, possibly based
on their pre-treatment covariates. Under a complete data model --- that is, a
causal model --- we assume that we have access to $n$ independent and
identically distributed (i.i.d.) copies of $X = (W, A, Y^{(0)}, Y^{(1)})$. Here,
$W$ is a random vector of covariates that includes the treatment--outcome
confounders, if any. $A$ is a random binary variable representing treatment
assignment; study units are said to be assigned to the treatment group when
$A=1$ and to the control group otherwise. $Y^{(0)}$ and $Y^{(1)}$ are potential
outcomes, continuous random variables representing the outcome under assignment
to the treatment and control groups, respectively.

## Causal Estimands

While we will never have access to copies of $X$, the complete data model allows
us to clearly define the causal estimands of interest. In particular, we will
consider differential variance parameters based on the potential outcome
variances, defined as $\sigma_c^2(a) = Var[Y^{(a)}]$. The potential outcome
variance under treatment is given by $\sigma_c^2(1)$, that under control is
given by $\sigma_c^2(0)$. Here, the subscript $c$ indicates that these are
parameters of a complete data model.

### Absolute Differential Variance

The first differential variance that we consider is the difference of the
potential outcomes' standard deviations. That is,
$$
\psi_c = \sigma_c(1) - \sigma_c(0) \;.
$$
This causal estimand contrasts the variability between groups on the absolute
scale. Positive values indicate that the treatment's potential outcomes are more
variable than the control's, negative values indicate the opposite, and a value
of zero indicates that the potential outcomes have identical variance.

### Relative Differential Variance

The second differential variance considered here is the ratio of the potential
outcomes' variances:
$$
\lambda_c = \frac{\sigma_c^2(1)}{\sigma_c^2(0)} \;.
$$
This causal estimand reports the groups variability on the relative scale.
Values larger than one correspond to the potential outcomes under treatment are
more variable, values below zero that the potential outcomes under control are
more variable, and a value of one indicates that the potential outcomes'
variances are identical.

## Uncovering Heterogeneous Treatment Effects

Now, if the treatment effect were homogeneous, $\psi_c$ would necessarily equal
zero and $\lambda_c$ would equal 1. This is because, in a homogeneous treatment
effect regime, the potential outcomes under treatment equal the potential
outcomes under control shifted by the ATE. By the linearity of expectation, the
variances of the potential outcomes are therefore identical.

Conversely, if $\psi_c \neq 0$ (or, equivalent, if $\lambda_c \neq 1$), the
treatment effect cannot be homogeneous. Causal inference about
$\psi_c$ and $\lambda_c$ therefore provides a straightforward test about
treatment effect heterogeneity. The formal theorem and accompanying proof of
this result are presented provided by @boileau2025.


## Observed Data and Identifiability Conditions

The differential variance estimands introduced above are defined in terms
of potential outcomes which are never available in practice. Luckily, we can
rely on standard --- but unverifiable --- causal identifiability assumptions to
perform inference about these causal estimands from observable data.

Under an observed data model, we have access to $n$ i.i.d. copies of random
observations $O = (W, A, Y)$, where $W$ and $A$ are as defined in $X$. Here, $Y$
corresponds to a random variable of the observation's continuous outcome. Then,
under the assumption of full exchangeability, positivity, and consistency,
we obtain for $a=0$ or $a=1$ the following results [@boileau2025]:
$$
\begin{split}
  \sigma_c^2(a)
  & = \mathbb{E}\left[\mathbb{E}[Y^2|W, A=a]\right] -
          \mathbb{E}\left[\mathbb{E}[Y|W,A=a]\right]^2 \\
  & = \sigma^2(a) \;.
\end{split}
$$
The assumption of full exchangeability assumes that $W$ contains all confounders
of the treatment--outcome relationship. The positive assumption requires that
study units have a non-zero probability of being assigned to either group.
Finally, the consistency assumptions states that the observed outcomes
correspond to the potential outcome associated with the study units' group
assignments. Of note, these identifiability conditions are satisfied by
construction in randomized controlled trials.

When these conditions are satisfied, we can in turn define parameters in the
observed data model that are equivalent to the differential variance estimands
previously defined in the complete data model. That is.
$$
\begin{split}
  \psi_c
  & = \sigma(1) - \sigma(0) \\
  & = \psi
\end{split}
$$
and
$$
\begin{split}
  \lambda_c
  & = \frac{\sigma^2(1)}{\sigma^2(0)} \\
  & = \lambda \;.
\end{split}
$$

We can therefore causal perform inference about these differential variance
parameters without requiring access to the potential outcomes, assuming the
causal identifiability conditions are satisfied. The `cmldiffvar` package
implements such procedures, which we briefly describe next.

# `cmldiffvar` in Action


## Unadjusted Estimators


### Estimators


### Example

```{r setup}
library(cmldiffvar)
```



## Adjusted Estimators


### Estimators


### Example


# Session Information

```{r session-information, echo=FALSE}
sessionInfo()
```

# References
