---
title: "Using the cmldiffvar Package"
author: "[Philippe Boileau](https://pboileau.ca)"
date: "`r Sys.Date()`"
bibliography: ../inst/references.bib
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{using-cmldiffvar}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Background

The average treatment effect (ATE) is a common causal estimand in causal
analyses. Defined as the difference of the mean potential outcomes, it
corresponds to the difference in average outcomes had all study units been
assigned to the treatment groups versus had they all been assigned to the
control group. While the ATE provides an important summary of the treatment
effect, it says nothing of effect heterogeneity. Even when the ATE indicates a
net positive effect of treatment on the average outcome, some study units may
experience better outcomes when assigned to the control. Decisions informed
solely by inference about the ATE might therefore produce unwelcome results;
effect heterogeneity, if it exists, must be accounted for.

The conditional ATE (CATE), another causal parameter, has gained popularity as
it can uncover and quantify this effect heterogeneity. As its name suggests, the
CATE reports the ATE of study units as a function of their pre-treatment
covariates. However, the CATE can only capture treatment effect heterogeneity if
the considered covariates are treatment effect modifiers. When the true effect
modifiers are unknown or they are missing from the data due to, for example, 
resource constraints or measurement issues, then the CATE will incorrectly
suggest that a treatment effect is homogeneous. Another inferential strategy
not subject to these limitations must supplement inference about the ATE.

We propose causal inference about the potential outcomes' variances to
accomplish just that. In particular, we have shown that tests about contrasts of
potential outcomes' variances, which we refer to differential variance, can be
used to formally test the homogeneous treatment effect assumption
[@boileau2025]. We briefly introduce this procedure in the next section.


# Differential Variance Inference

## Complete Data Model

We assume that data are generated according to a parallel group study design
wherein study units are randomly assigned to one of two groups, possibly based
on their pre-treatment covariates. Under a complete data model --- that is, a
causal model --- we assume that we have access to $n$ independent and
identically distributed (i.i.d.) copies of $X = (W, A, Y^{(0)}, Y^{(1)})$. Here,
$W$ is a random vector of covariates that includes the treatment--outcome
confounders, if any. $A$ is a random binary variable representing treatment
assignment; study units are said to be assigned to the treatment group when
$A=1$ and to the control group otherwise. $Y^{(0)}$ and $Y^{(1)}$ are potential
outcomes, continuous random variables representing the outcome under assignment
to the treatment and control groups, respectively.

## Causal Estimands

While we will never have access to copies of $X$, the complete data model allows
us to clearly define the causal estimands of interest. In particular, we will
consider differential variance parameters based on the potential outcome
variances, defined as $\sigma_c^2(a) = Var[Y^{(a)}]$. The potential outcome
variance under treatment is given by $\sigma_c^2(1)$, that under control is
given by $\sigma_c^2(0)$. Here, the subscript $c$ indicates that these are
parameters of a complete data model.

### Absolute Differential Variance

The first differential variance that we consider is the difference of the
potential outcomes' standard deviations. That is,
$$
\psi_c = \sqrt{\sigma_c^2(1)} - \sqrt{\sigma_c^2(0)} \;.
$$
This causal estimand contrasts the variability between groups on the absolute
scale. Positive values indicate that the treatment's potential outcomes are more
variable than the control's, negative values indicate the opposite, and a value
of zero indicates that the potential outcomes have identical variance.

### Relative Differential Variance

The second differential variance considered here is the ratio of the potential
outcomes' variances:
$$
\lambda_c = \frac{\sigma_c^2(1)}{\sigma_c^2(0)} \;.
$$
This causal estimand reports the groups variability on the relative scale.
Values larger than one correspond to the potential outcomes under treatment are
more variable, values below zero that the potential outcomes under control are
more variable, and a value of one indicates that the potential outcomes'
variances are identical.

## Uncovering Heterogeneous Treatment Effects

Now, if the treatment effect were homogeneous, $\psi_c$ would necessarily equal
zero and $\lambda_c$ would equal 1. This is because, in a homogeneous treatment
effect regime, the potential outcomes under treatment equal the potential
outcomes under control shifted by the ATE. By the linearity of expectation, the
variances of the potential outcomes are therefore identical.

Conversely, if $\psi_c \neq 0$ (or, equivalent, if $\lambda_c \neq 1$), the
treatment effect cannot be homogeneous. Causal inference about
$\psi_c$ and $\lambda_c$ therefore provides a straightforward test about
treatment effect heterogeneity. The formal theorem and accompanying proof of
this result are presented provided by @boileau2025.


## Observed Data and Identifiability Conditions

The differential variance estimands introduced above are defined in terms
of potential outcomes which are never available in practice. Luckily, we can
rely on standard --- but unverifiable --- causal identifiability assumptions to
perform inference about these causal estimands from observable data.

Under an observed data model, we have access to $n$ i.i.d. copies of random
observations $O = (W, A, Y)$, where $W$ and $A$ are as defined in $X$. Here, $Y$
corresponds to a random variable of the observation's continuous outcome. Then,
under the assumption of full exchangeability, positivity, and consistency,
we obtain for $a=0$ or $a=1$ the following results [@boileau2025]:
$$
\begin{split}
  \sigma_c^2(a)
  & = \mathbb{E}\left[\mathbb{E}[Y^2|W, A=a]\right] -
          \mathbb{E}\left[\mathbb{E}[Y|W,A=a]\right]^2 \\
  & = \sigma^2(a) \;.
\end{split}
$$
The assumption of full exchangeability assumes that $W$ contains all confounders
of the treatment--outcome relationship. The positive assumption requires that
study units have a non-zero probability of being assigned to either group.
Finally, the consistency assumptions states that the observed outcomes
correspond to the potential outcome associated with the study units' group
assignments. Of note, these identifiability conditions are satisfied by
construction in randomized controlled trials.

When these conditions are satisfied, we can in turn define parameters in the
observed data model that are equivalent to the differential variance estimands
previously defined in the complete data model. That is.
$$
\begin{split}
  \psi_c
  & = \sqrt{\sigma^2(1)} - \sqrt{\sigma^2(0)} \\
  & = \psi
\end{split}
$$
and
$$
\begin{split}
  \lambda_c
  & = \frac{\sigma^2(1)}{\sigma^2(0)} \\
  & = \lambda \;.
\end{split}
$$

We can therefore causal perform inference about these differential variance
parameters without requiring access to the potential outcomes, assuming the
causal identifiability conditions are satisfied. The `cmldiffvar` package
implements such procedures, which we briefly describe next.

# `cmldiffvar` in Action

The `cmldiffvar` package implements two types of estimators for the absolute and
relative differential variance: unadjusted and adjusted estimators. The former
is appropriate for randomized experiments in which there are no confounders of
the treatment--outcome relationship. The latter should be used in analyses of
observational study data to account for confounding variables. Adjusted
estimators can also be used in randomized experiments to adjust for
pre-treatment covariates that explain variability in the outcome variable,
potentially improving upon the precision of the unadjusted estimator.

## Unadjusted Estimators

The unadjusted estimators of the absolute and relative differential variance are
defined as the difference of the groups' outcomes' standard deviations and the
ratio of the groups' outcomes' variances, respectively. That is,
$$
\hat\psi = \sqrt{\hat\sigma^2(1)} - \sqrt{\hat\sigma^2(0)}
$$
and 
$$
\hat\lambda = \frac{\hat\sigma^2(1)}{\hat\sigma^2(0)} \;,
$$
where
$$
\hat\sigma^2(a) = \widehat{\text{Var}}[Y|A=a] \;,
$$
the group-specific sample variance.

The unadjusted absolute and relative differential variance estimators are
implemented in the `unadjdiffvar()` function. The absolute differential variance
estimator is called by default; the relative differential variance estimator is
called by setting the `estimand_type` function to `"relative"`.

These estimators' asymptotic sampling distributions are normal with means equal
to their respective target estimand and with variance given by the variances of
their respective influence function. Their influence functions can therefore be
used to construct Wald-type 95% confidence intervals, which in turn permit
hypothesis testing. In particular, these estimators can be used to perform
hypothesis tests about the homogeneous treatment effect assumption. Under the
null hypothesis, $\psi = 0$ and $\lambda = 1$. Under the alternative,
$\psi \neq 0$ and $\lambda \neq 1$.

### Example

We estimate the relative differential variance from a simulated randomized
controlled trial using the `unadjdiffvar()` function. We must specify the
estimand using the `estimand_type` argument, and provide the variable names
of the treatment, outcome, and propensity score variable names in the sample
data set. If the propensity score variable name is not provided to
`unadjdiffvar()`, they will be estimated from the provided data set as the
proportion of study units assigned to the treatment group.

The variances of the potential outcomes under the treatment and control group
assignments are $1.44$ and $1$, respectively. The relative differential variance
parameter therefore has a value of $\lambda_c = 1.44$, indicating that the
treatment effect is heterogeneous.

```{r unadjusted-estimator}
# load the required libraries
library(cmldiffvar)
library(dplyr)

# set the seed for reproducibility
set.seed(3976468)

# simulate randomized controlled trial population
n_pop <- 100000
propensity_score <- 0.5
propensity_score_vec <- rep(propensity_score, n_pop)
treatment_vec <- rbinom(n_pop, 1, propensity_score)
outcome_treatment_vec <- rnorm(n = n_pop, mean = 3, sd = 1.2)
outcome_control_vec <- rnorm(n = n_pop, mean = 1, sd = 1)
outcome_vec <- treatment_vec * outcome_treatment_vec +
  (1 - treatment_vec) * outcome_control_vec
population_tbl <- tibble(
  propensity_score = propensity_score_vec,
  treatment = treatment_vec,
  outcome = outcome_vec
)

# randomly sample from the population
sample_tbl <- slice_sample(population_tbl, n = 500)

# estimate the relative differential variance and test the homogeneous treatment
# assumption
rel_diff_var_est <- sample_tbl |>
  unadjdiffvar(
    estimand_type = "relative",
    treatment_var_name = "treatment",
    propensity_score_var_name = "propensity_score",
    outcome_var_name = "outcome"
  )

rel_diff_var_est
```

The unadjusted estimator produces an estimate of
`r round(rel_diff_var_est$estimate, 3)`, suggesting that the potential outcomes
under the treatment condition are more variable than the potential outcomes
under control. The 95% confidence interval does not cover $1$, providing
evidence of a statistically significant heterogeneous treatment effect. Indeed,
the p-value of the associated test is `r round(rel_diff_var_est$p_value, 3)`
such that the homogeneous treatment effect hypothesis is rejected.

## Adjusted Estimators

The `cmldiffvar` package implements two adjusted estimators for each
differential variance parameter: a one-step estimator and a targeted maximum
likelihood estimator (TMLE). These estimators rely on the estimation of three
nuisance parameters:

1) the propensity score: $g(W) = P(A=1|W)$,
2) the outcome regression: $\bar{Q}(W, A) = E[Y|W, A]$, and 
3) the squared outcome regression: $\bar{Q}^2(W, A) = E[Y^2|W, A]$ .

These adjusted estimators are doubly robust. That is, if $g(W)$ or both
$\bar{Q}(W, A)$ and $\bar{Q}^2(W, A)$ are consistently estimated, then the
adjusted estimators are consistent as well. Further, if the nuisance estimators
of $g(W)$, $\bar{Q}(W, A)$, and $\bar{Q}^2(W, A)$ are consistent and converge at
a rate of $n^{1/4}$ under the $L_2$ norm, then these estimators are regular and
asymptotically linearly. This implies that these differential variance parameter
estimators' sampling distributions are asymptotically normal about the true
differential variance parameters with variance given by the parameters'
efficient influence functions. As with the unadjusted estimators, 95% Wald-type
confidence intervals can in turn be constructed, and hypotheses about treatment
effect heterogeneity can be performed. Additional details are provided in
@boileau2025.

Now, this package relies on Super Learners to flexibly estimate $g(W)$,
$\bar{Q}(W, A)$, and $\bar{Q}^2(W, A)$ [@laanSuperLearner2007]. In particular,
the `cmldiffvar` package relies on the
[`SuperLearner`](https://cran.r-project.org/web/packages/SuperLearner/index.html)
R package implementation of this ensemble learning framework. By default, a
generalized linear model, multivariate adaptive regression splines, and a Random
Forest are used as base learners in the Super Learner estimator of the
propensity score and the outcome regression. A generalized quadratic linear
model and a Random Forest make up the base learners of the Super Learner
estimator for the squared outcome regression. Of course, an alternative set of
base learners used for each estimator can be specified in `cmldiffvar()`
function.

Finally, we note that these adjusted estimators are guaranteed to be regular and
asymptotically linear when applied to data generated in a randomized controlled
trial. The known propensity scores, corresponding to study units arm
randomization probabilities, can be provided to the `cmldiffvar()` function
using its `propensity_score_var_name` argument.

### Example

We estimate the absolute differential variance from a random of a simulated
observational study population provided with the `cmldiffvar` package using the
TMLE implemented in the `cmldiffvar()` function. Similar to the unadjusted
estimator, we must provide the names of the treatment and outcome variables in
the sample data set. The treatment--outcome relationship is confounded by a
single variable called `"confounder"`, which we adjust for using the
`propensity_score_adj_var_names` and `cond_exp_outcome_adj_var_names` arguments.
Note that the adjustment sets for the propensity score and conditional (squared)
outcome regressions are specified separately to allow for increased flexibility
when estimating nuisance parameters. Finally, we rely on the default Super
Learners to estimate the nuisance parameters.

The variance of the potential outcomes under treatment and control are $10$ and
$2$, respectively. The absolute differential variance estimand therefore has a
value of $1.75$.

```{r adjusted-estimator}
# load the required libraries
library(cmldiffvar)
library(dplyr)
library(SuperLearner)

# set the seed for reproducibility
set.seed(86523)

# randomly sample from the provided observational study population
sample_tbl <- slice_sample(toy_population_tbl, n = 500)

# estimate the relative differential variance and test the homogeneous treatment
# assumption
abs_diff_var_est <- sample_tbl |>
  cmldiffvar(
    estimator_type = "tmle",
    treatment_var_name = "treatment",
    outcome_var_name = "outcome",
    propensity_score_adj_var_names = "confounder",
    cond_exp_outcome_adj_var_names = "confounder" 
  )

abs_diff_var_est
```

The TMLE produces an estimate of `r round(abs_diff_var_est$estimate, 3)`,
indicating that the standard deviation of the potential outcomes under treatment
is greater than that of the potential outcomes under control. Further, the 95%
confidence interval and the p-value of `r round(abs_diff_var_est$p_value, 3)`
suggest that we reject the null hypothesis of a homogeneous treatment effect.
Near identical results are obtained when using a one-step estimator instead,
which can be specified by setting the `estimator_type` argument to `"one-step"`.

# Session Information

```{r session-information, echo=FALSE}
sessionInfo()
```

# References
